---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- use devtools::build_readme() -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# omopcept

<!-- badges: start -->
<!-- badges: end -->

omopcept provides access to a subset of **OMOP** con**CEPT**s (without the cons!) and flexible tidyverse compatible R functions for querying.    
     
On initial use it downloads a [parquet](https://parquet.apache.org/) file containing the concepts to a local package cache where it can be accessed in future sessions. On use it opens the file (with omop_concept()) but doesn't read in all the data to save time. The [arrow R package](https://arrow.apache.org/docs/r/index.html) allows parquet files to be opened and queried in dplyr pipelines without having to read in all the data.  e.g. the code below will return just the top rows of the concept table.    

``` r

omop_concept() |> 
  head() |> 
  dplyr::collect()

```

## Installation

Install the development version of omopcept with:

``` r
# install.packages("remotes")

remotes::install_github("andysouth/omopcept")

```

## Main functions

```omop_names() onames()```     search concepts by parts of names     
     
```omop_id() oid()```           search for a concept_id     
     
```omop_join_names()```         join names onto a table with an id column     

```omop_concept()```            return reference to the concept table that can be used in dplyr pipelines         

## OMOP background

The [OMOP Common Data Model](https://ohdsi.github.io/CommonDataModel/) is an open standard for health data. "[It is] designed to standardize the structure and content of observational data and to enable efficient analyses that can produce reliable evidence".    
    
OMOP is maintained by OHDSI (pronounced "Odyssey"). "The Observational Health Data Sciences and Informatics  program is a multi-stakeholder, interdisciplinary collaborative that strives to improve medical decision making and bring better health outcomes to patients around the world."    
OMOP concepts can be searched and downloaded from [Athena – the  OHDSI vocabularies repository](https://athena.ohdsi.org). This package provides R tools to interact with the concepts in a more reproducible way.   

## Concept data

OMOP vocab data downloaded from Athena includes a table called CONCEPT.csv that we saved in parquet format for use in this package.     
     
omopcept downloads a selection of vocabularies and stores locally the first time you use it.


```{r concept table, echo=FALSE}

library(omopcept)

fieldnames <- omop_concept() |> 
  head() |> 
  names()

df1 <- data.frame(fields=fieldnames,
                  about=c("unique id","descriptive name","e.g. drug, measurement","e.g. LOINC, SNOMED","e.g. Clinical Observation, Organism","standard or not","source code","","",""),
                  query_arguments=c("c_ids","pattern","d_ids","v_ids","cc_ids","standard","","","","")
                  )

knitr::kable(df1)
```

## String search in concept_name field

```{r omop_names, echo=TRUE}

omop_names("chemotherapy", v_ids="LOINC")

omop_names("chemotherapy", v_ids=c("LOINC","SNOMED"), d_ids=c("Observation","Procedure"))

```

## Join OMOP names onto a dataframe containing concept ids in a column called *concept_id

Helps to interpret OMOP data.

```{r omop_join_name, echo=TRUE}


data.frame(concept_id=(c(3571338L,4002075L))) |> 
  omop_join_name()
 

data.frame(drug_concept_id=(c(4000794L,4002592L))) |> 
  omop_join_name(namestart="drug")


```


## Vocabularies included

The vocabularies are a default download from Athena with a few extra vocabs added. 
Later we may offer option to add other vocabularies.


### Numbers of concepts in the package by domain and vocabulary

```{r conceptplot, echo = TRUE, fig.height=12}
library(dplyr)
library(ggplot2)
library(forcats)

concept_summary <- 
  omop_concept() |>
  count(vocabulary_id, sort=TRUE) |> 
  collect()

ggplot(concept_summary,aes(y=reorder(vocabulary_id,n),x=n,col=vocabulary_id)) +
  geom_point() +
  labs(y = "vocabulary_id") +
  guides(col="none") +
  theme_minimal()


```

