---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- use devtools::build_readme() -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# omopvocabr

<!-- badges: start -->
<!-- badges: end -->

omopvocabr provides access to a subset of OMOP Common Data Model medical vocabularies and flexible tidyverse compatible R functions for querying.

## Installation

Install the development version of omopvocabr from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("andysouth/omopvocabr")
```

## Example showing what vocabularies are included

Later there may be options to include more.

```{r example}

library(omopvocabr)
library(dplyr)

## showing what vocabs are included
concept |> count(vocabulary_id)

```

### Numbers of concepts in the package by domain and vocabulary

```{r conceptplot, echo = TRUE}
library(ggplot2)
library(forcats)

ggplot(concept, aes(y=fct_rev(fct_infreq(domain_id)), 
                    fill=vocabulary_id)) +
  geom_bar() +
  labs(y = "domain_id") +
  theme_minimal()
```

