---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- use devtools::build_readme() -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# omopcepts

<!-- badges: start -->
<!-- badges: end -->

omopcepts provides access to a subset of **OMOP** con**cepts** and flexible tidyverse compatible R functions for querying.    

The [OMOP Common Data Model](https://ohdsi.github.io/CommonDataModel/) is an open standard for health data. "[It is] designed to standardize the structure and content of observational data and to enable efficient analyses that can produce reliable evidence".

## Installation

Install the development version of omopcepts with:

``` r
# install.packages("remotes")
remotes::install_github("andysouth/omopcepts")
```

## Join OMOP names onto a dataframe containing *concept_id

Helps to interpret OMOP data.

```{r join_omop_name, echo=TRUE}

library(omopcepts)

df1 <- data.frame(concept_id=(c(3571338L,4002075L)))

join_omop_name(df1)

```

## String search in concept_name field

```{r concept_names, echo=TRUE}

concept_names("chemotherapy", v_ids="LOINC")

```

## Vocabularies included

Initially just a few vocabularies are included to keep the size of the data file down. Later we may offer option to add other vocabularies.

```{r example}

library(omopcepts)
library(dplyr)

## showing what vocabs are included
concept |> count(vocabulary_id)

```

### Numbers of concepts in the package by domain and vocabulary

```{r conceptplot, echo = TRUE}
library(ggplot2)
library(forcats)

ggplot(concept, aes(y=fct_rev(fct_infreq(domain_id)), 
                    fill=vocabulary_id)) +
  geom_bar() +
  labs(y = "domain_id") +
  theme_minimal()
```

