---
title: "Background to the data structures used for plotting networks and how these are derived from OMOP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data-structures-for-plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(omopcept)
```

`omopcept` relies on the R packages [`ggraph`](https://ggraph.data-imaginist.com/index.html) and [`tidygraph`](https://tidygraph.data-imaginist.com/) for plotting omop relationships in the `omop_graph()` function.

`omop_graph()` | is divided into two parts 
-------------- | -------------
`omop_graph_calc()` | calculates a plotting data structure from OMOP vocabulary data 
`omop_graph_vis()`  | visualises a data structure created above


`omop_graph_calc()` produces two tables one for nodes (points) and one for edges (connections between points). A very simple example could look like this.

nodes
from to

edges

