---
title: "Getting drug classes from the OMOP hierarchy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting drug classes from the hierarchy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(omopcept)
library(CodelistGenerator)
library(dplyr)
```

**DRAFT**   

## TLDR

omopcept has recently gained a function `omop_drug_lookup_create()` to create a lookup table from drug concept IDs to drug classes in the ATC classification. To create a lookup table for all RxNorm Extension Ingredients :

```
drug_lookup = omop_drug_lookup_create()
```

## Use case

Researchers want to be able to identify drug classes from an omop extract that has more specific drug names.

Maybe 
1. antibiotics
1. broad spectrum antibiotics
1. specific antibiotic classes

## ATC a WHO drug classification

[ATC](https://atcddd.fhi.no/atc/structure_and_principles/) the Anatomical Therapeutic Chemical Classification System maintained by WHO and the Defined Daily Dose (DDD) as a measuring unit "have become the gold standard for international drug utilization monitoring and research".

In ATC, active substances are classified in a hierarchy with five levels.  The system has fourteen main anatomical/pharmacological groups or 1st levels.  Each ATC main group is divided into 2nd levels which could be either pharmacological or therapeutic groups.  The 3rd and 4th levels are chemical, pharmacological or therapeutic subgroups and the 5th level is the chemical substance.

## ATC in OMOP via RxNorm Extension & RxNorm

In OMOP each drug concept ID in the vocabularies `RxNorm Extension` & `RxNorm` has ancestor concepts in the different levels of the ATC hierarchy. The new omopcept function `omop_drug_lookup_create()` creates a lookup table from drug concept IDs to drug classes in the ATC classification.



## Does anything else already do this ?

I couldn't quite get anything to provide a comprehensive list of drug classes.

Seems that the [CodelistGenerator package](https://github.com/darwin-eu/CodelistGenerator) does things that could solve this.

[CodelistGenerator::getATCCodes](https://github.com/darwin-eu/CodelistGenerator/blob/main/R/drugCodes.R) returns descendants of ATC classes.

@param level ATC level. Can be one or more of "ATC 1st", "ATC 2nd", "ATC 3rd", "ATC 4th", and "ATC 5th"
@param name ATC name of interest. For example, c("Dermatologicals", "Nervous System"), would result in a list of length two with the descendant concepts for these two particular ATC groups.

```{r CodeListGenerator getATCCodes, eval=FALSE}

cdm <- mockVocabRef()
atc1 <- getATCCodes(cdm = cdm, level = "ATC 1st")
# atc2 <- getATCCodes(cdm = cdm, level = "ATC 2nd")
# Error in getATCCodes(cdm = cdm, level = "ATC 2nd") : 2 assertions failed:
#  * - No matching ATC codes found
#  * Variable 'atcCheck': Must be TRUE.

```

But these only show results that are in a CDM. I'd like comprehensive results to help me make sure its doing what I want before I apply it to any data.


```{r omopcept subset ATC levels from concepts table, eval=FALSE}

# OLD exploration
# eval=FALSE because much of this replaced by omop_drug_lookup_create()

#300k rows
rxn <- omopcept::omop_names("", v="RxNorm")

omopfreqconceptclass(rxn)

# doesn't seem to have ATC classes I want ??

#    concept_class_id         n
#    <chr>                <int>
#  1 Clinical Drug        53083
#  2 Branded Drug         39083
#  3 Clinical Drug Comp   38474
#  4 Branded Drug Comp    36748
#  5 Branded Drug Form    26595
#  6 Branded Dose Group   23388
#  7 Clinical Drug Form   20773
#  8 Brand Name           19469
#  9 Clinical Dose Group  17741
# 10 Ingredient           15604
# 11 Multiple Ingredients  4469
# 12 Quant Clinical Drug   3860
# 13 Precise Ingredient    3476
# 14 Quant Branded Drug    3215
# 15 Branded Pack          1308
# 16 Clinical Pack         1175
# 17 Dose Form              201
# 18 Dose Form Group         47

# Aha, by looking at code for CodeListGenerator I see that the ATC classes are not in RxNorm but in a vocab called ATC

#6740 rows
atc <- omopcept::omop_names("", v="ATC")

omopfreqconceptclass(atc)
# 1 ATC 5th           5452
# 2 ATC 4th            911
# 3 ATC 3rd            269
# 4 ATC 2nd             94
# 5 ATC 1st             14

# TO be able to get ATC classes for any drug concept
# take the ancestor table
# filter all ancestor_concept_id that are in vocab ATC
# can then use filter any descendant_concept_id that appears in user data
# and should end up with vector of ATC classes at the different levels

#9 million ATC descendants !!

atc_descendants <- omop_concept_ancestor() |> 
  #ideally would do before collect
  #but get Error in df[[id_col_name]] <- as.integer(df[[id_col_name]]) : cannot add bindings to a locked environment
  #that I can probably fix in omopcept
  #collect() |> 
  omop_join_name(namestart = "ancestor", columns = c("concept_name","vocabulary_id","concept_class_id")) |> 
  #renaming of joined columns to differentiate ancestor & descendant
  rename(ancestor_vocabulary_id = vocabulary_id) |> 
  rename(ancestor_concept_class_id = concept_class_id) |>  
  omop_join_name(namestart = "descendant", columns = c("concept_name","vocabulary_id","concept_class_id")) |> 
  rename(descendant_vocabulary_id = vocabulary_id) |> 
  rename(descendant_concept_class_id = concept_class_id) |> 
  collect() |>   
  #head(100) |> 
  filter(ancestor_vocabulary_id=="ATC") 

atc_descendants |> count(descendant_vocabulary_id, sort=TRUE)
# 1 RxNorm Extension         8732808
# 2 RxNorm                    819887
# 3 ATC                        30849
# 4 HCPCS                        140
# 5 HemOnc                        66

atc_descendants |> count(ancestor_concept_class_id, sort=TRUE)

#8.7m rows
atc_rxnormext_descendants <- atc_descendants |> 
  filter(descendant_vocabulary_id == "RxNorm Extension") |> 
  select(ATC_concept_id = ancestor_concept_id,
         ATC_concept_name = ancestor_concept_name,
         drug_concept_id = descendant_concept_id,
         drug_concept_name = descendant_concept_name,
         ATC_level = ancestor_concept_class_id,
         #this is RxNormExtension concept class e.g. Branded Drug etc.
         #probably don't need because it will be in drug_exposure table
         #but may be useful here during development
         concept_class_id = descendant_concept_class_id
         ) |> 
   #extract numeric part of the ATC level
   mutate(ATC_level = str_sub(ATC_level,5,5))

atc_rxnormext_descendants |> count(concept_class_id, sort=TRUE)

#  1 Marketed Product    2692126
#  2 Branded Drug Box    1084960
#  3 Branded Drug         856819
#  4 Branded Drug Comp    710365
#  5 Branded Drug Form    569727
#  6 Clinical Drug Box    532821
#  7 Quant Branded Drug   467679
#  8 Quant Branded Box    411968
#  9 Quant Clinical Drug  346667
# 10 Clinical Drug        333783
# 11 Quant Clinical Box   324265
# 12 Clinical Drug Comp   200522
# 13 Clinical Drug Form   151217
# 14 Branded Pack          16291
# 15 Clinical Pack         12935
# 16 Ingredient             8911
# 17 Branded Pack Box       6604
# 18 Clinical Pack Box      5148

#SOMETHING NOT RIGHT ? SEEMS CAN'T FIND MANY DRUGS IN HERE

#TODO check in one of UCLH extracts what the values of RxNormExt concept_class_id are
#e.g. is it just Ingredient ?

atc_rxnormext_ingredient <- atc_rxnormext_descendants |> 
  filter(concept_class_id %in% c("Ingredient"))

atc2ingredients <- atc_rxnormext_ingredient |> 
  filter(ATC_level %in% c("ATC 1st"))

atc_rxnormext_clinicaldrug <- atc_rxnormext_descendants |> 
  filter(concept_class_id %in% c("Clinical Drug"))

atc_tmp <- atc_rxnormext_descendants |> head(1000)

atc_from_rxnormext_ingredient <- atc_rxnormext_ingredient |> 
#atc_from_rxnormext_wide <- atc_rxnormext_descendants |> 
#atc_from_rxnormext_wide <- atc_tmp |> 
  pivot_wider(names_from = ATC_level, 
              values_from = c(ATC_concept_name, ATC_concept_id), 
              id_cols=c(drug_concept_name, drug_concept_id))

# Warning message:
# Values from `ATC_concept_id` and `ATC_concept_name` are not uniquely identified; output will contain list-cols.
# Use the following dplyr code to identify duplicates.
# 800k rows !
#dup <- atc_rxnormext_descendants |>
dup <- atc_rxnormext_ingredient |>  
  dplyr::summarise(n = dplyr::n(), .by = c(drug_concept_name, drug_concept_id, ATC_level)) |>
  dplyr::filter(n > 1L)

# SO a drug can be in more than one ATC class

# lets have a look at a single drug example

drugname <- "biapenem"
drugname <- "Bromisoval" #apparently has 37 duplicates !!
drugname <- "Azidamfenicol" #10 duplicates

oneingredient <- atc_rxnormext_descendants |> 
  filter(drug_concept_name == drugname) |> 
  arrange(ATC_level)

#biapenem has single entry per ATC level
#   ATC_concept_id ATC_concept_name                 drug_concept_id drug_concept_name ATC_level concept_class_id
# 1       21602795 ANTIINFECTIVES FOR SYSTEMIC USE         35198093 biapenem          ATC 1st   Ingredient      
# 2       21602796 ANTIBACTERIALS FOR SYSTEMIC USE         35198093 biapenem          ATC 2nd   Ingredient      
# 3       21602868 OTHER BETA-LACTAM ANTIBACTERIALS        35198093 biapenem          ATC 3rd   Ingredient      
# 4       21602920 Carbapenems                             35198093 biapenem          ATC 4th   Ingredient      
# 5       21602924 biapenem; parenteral                    35198093 biapenem          ATC 5th   Ingredient  

#but e.g. Azidamfenicol has multiple entries in all ATC levels
#suggests that maybe we need multiple rows per drug (i.e. keep data long)
#perhaps function can offer option for short & long ?
#OR maybe start by producing a long lookup table because we are close to that already
#

#even Bromisoval, most duplicates are in level 5
#oneingredient |> count(ATC_level, sort=TRUE)
# 1 ATC 5th      37
# 2 ATC 4th       6
# 3 ATC 3rd       4
# 4 ATC 1st       1
# 5 ATC 2nd       1

#TEMPTED to try to plot relationships of ATC levels
#probably 1-3


#I think we may want ATC (3rd level, pharmacological subgroup)
atc3_rxnormext_descendants <- atc_descendants |> 
  filter(descendant_vocabulary_id == "RxNorm Extension") |> 
  filter(ancestor_concept_class_id == "ATC 3rd")

atc2_rxnormext_descendants <- atc_descendants |> 
  filter(descendant_vocabulary_id == "RxNorm Extension") |> 
  filter(ancestor_concept_class_id == "ATC 2nd")

atc1_rxnormext_descendants <- atc_descendants |> 
  filter(descendant_vocabulary_id == "RxNorm Extension") |> 
  filter(ancestor_concept_class_id == "ATC 1st")

#234 rows at ATC3
freq_atc3 <- atc3_rxnormext_descendants |> count(ancestor_concept_name, sort=TRUE)
#90 rows at ATC2
freq_atc2 <- atc2_rxnormext_descendants |> count(ancestor_concept_name, sort=TRUE)
#14 rows at ATC1
freq_atc1 <- atc1_rxnormext_descendants |> count(ancestor_concept_name, sort=TRUE)

cyto <- atc3_rxnormext_descendants |> 
  filter(ancestor_concept_name == "CYTOTOXIC ANTIBIOTICS AND RELATED SUBSTANCES")

#amoxicillin etc.
beta <- atc3_rxnormext_descendants |> 
  filter(ancestor_concept_name == "BETA-LACTAM ANTIBACTERIALS, PENICILLINS")

#OTHER BETA-LACTAM ANTIBACTERIALS


#find what level2 is for level3= "BETA-LACTAM ANTIBACTERIALS, PENICILLINS"
beta_atc2 <- atc_descendants |> 
  filter(ancestor_concept_class_id == "ATC 2nd") |> 
  filter(descendant_concept_name == "BETA-LACTAM ANTIBACTERIALS, PENICILLINS")  

anti_atc3 <- atc_descendants |> 
  filter(ancestor_concept_name == "ANTIBACTERIALS FOR SYSTEMIC USE" ) |> 
  filter(descendant_concept_class_id == "ATC 3rd")

anti_atc4 <- atc_descendants |> 
  filter(ancestor_concept_name == "ANTIBACTERIALS FOR SYSTEMIC USE" ) |> 
  filter(descendant_concept_class_id == "ATC 4th")

beta_atc1 <- atc_descendants |> 
  filter(ancestor_concept_class_id == "ATC 1st") |> 
  filter(descendant_concept_name == "BETA-LACTAM ANTIBACTERIALS, PENICILLINS") 

a3_and_4 <- bind_rows(anti_atc3, anti_atc4)

#SO WHAT WE COULD DO is create a post-processing script that will add columns to the omop extracts for ATC2,3,& 4 likely to be most useful
#TODO
#write a function addATC_to_drug_concepts()
#want to add 1 column for each ATC level
#probably by filtering all ATC levels & then doing pivot_wider()

```

