---
title: "getting-omop-concept-data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-omop-concept-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(omopcept)
```


`omopcept` requires OMOP concept tables.

In default operation omopcept will download these concept tables from a temporary online filestore we have created.

You can also download concept tables from [Athena](https://athena.ohdsi.org/) and then use `omopcept::omop_vocabs_preprocess()` to save these in the `omopcept` package. Note that package results will depend on the subset of vocabularies that you have selected from the Ahena download page.


### Details of vocab storage

`omopcept` by default downloads a selection of vocabularies and stores them locally the first time you use it (in the recommended data location for R packages). The download does not need to be repeated unless you update the package. Vocabularies are stored as [parquet](https://parquet.apache.org/) files that can be queried in a memory-efficient manner without having to first read the data in to memory.

`omopcept::omop_vocab_table_save()` is used to save parquet files into the package that we have stored online. You could also use `omop_vocab_table_save(from=*)` to save vocab parquet files from your own online location.
