---
title: "Exploring OMOP hierarchy for blood counts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring OMOP hierarchy for blood counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup, include = FALSE}
library(omopcept)
library(stringr)
library(dplyr)
library(knitr) #for kable
```

## Use case

A need to map blood count components from two different source systems to OMOP. There are different potential OMOP concept ids. How to choose between them ? How to deal with a situation where different institutions end up with differeny OMOP concept ids for essentially the same measure.


### Leukocytes

Starting with a single concept id `3010813` (chosen in a mapping process) for the number of Leukocytes in a volume of blood. 

\small
```{r leukocytes, include=TRUE}
lvbg <- omop_id(3010813)

lvbg |> 
  select(-valid_start_date,-valid_end_date,-invalid_reason) |> 
  kable()
```
\normalsize

Using omopcept to find the immediate relations of "Leukocytes [#/volume] in Blood"

\small
```{r leukocyte volume relations}

leuk_r <- omop_relations(3010813)

leuk_r |> arrange(relationship_id) |>
  select(relationship_id,concept_name_2,domain_id,concept_class_id) |> 
  kable()
```
\normalsize

Of these relations `Leukocytes | Blood | Hematology and Cell counts` looks like it may be useful, it has a `concept_class_id` of `LOINC Hierarchy` and itself has the following relations.

\small
```{r leukocyte hierarchy relations}

lbhcc_r <- omop_relations(37043992) |> 
  arrange(relationship_id) 

lbhcc_r |>
  select(relationship_id,concept_name_2,domain_id,concept_class_id) |> 
  kable()

lbhcc_r |> 
  omop_graph( graphtitle = "", saveplot = FALSE )

```
\normalsize

Can we find similar hierarchy parents for other blood count concepts ?

\small
```{r loinc hierarchy for other blood counts}

#new fixed=TRUE arg for omop_names() allows this 
bhccounts <- omop_names("Blood | Hematology and Cell counts", fixed=TRUE) |> 
  #returning 452 concepts
  #removing those for Fetus
  filter(!str_detect(concept_name,"Fetus"))
  #now down to 417 concepts
```
\normalsize



See https://github.com/OHDSI/Vocabulary-v5.0/issues/301
which has some info about this
